Visualizing probability distributions across bivariate nonlinear temporal granularities

### 2020-02-24 main.Rmd
data structure introduction changed

Effective exploration or good visualization require good data structures. Commonly, simple sequences of time value pairs $(<t_0, v_0> \dots, <t_n, v_n>)$ are the basis of analysis and visualization. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of granularities. Data can then be consolidated for different levels of granularity enabling statistical summaries of values along granularities. Since, we are interested in detection of unknown periodic behavior of data, it makes sense to support the detection of patterns by obtaining statistical summaries across cyclic time granularities. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, the big problem is broken down into smaller pieces by focusing on two cyclic granularities at a time.

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, which will essentially mean that the measured variable can not be explored for all combinations of the categories/levels of the cyclic granularities. To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy.
These are called "structurally" empty combinations. Let us take a specific example, where $C_1$ be day-of-month and $C_2$ be week-of-month. Here $C_1$ can have 31 levels/categories while $C_2$ can have 5 categories. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of day-of-month and week-of-month. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc.This is also intuitive since the first day of the month can never correspond to fifth week of the month. These are structurally empty sets in that it is impossible for them to have any observations.

@@

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Again, let us consider a specific example to illustrate this. Let $C_1$ be day-of-week and $C_2$ be WorkingDay/NonWorkingDay. Here $C_1$ can have 7 levels while $C_2$ can have 2 levels. So there are 14 sets $S_{ij}$ corresponding to the possible combinations of $C_1$ and $C_2$. While potentially all of these can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these combinations will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ is day-of-month. Then the days denoting weekends in a month would not correspond to any Business days and would have missing observations due to the way the granularities are constructed. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does. Hence, they are referred to as "build-based" empty combinations.

An example when there will be no empty combinations could be where $C_1$ maps index set to day-of-week and $C_2$ maps index set to month-of-year. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of day-of-week and month-of-year. All of these are non-empty because every day-of-week can occur in every month.

@@

## Measured variable {#sec:measured-variable}

@@

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. With huge amount of data being available, mean, median or any one summary statistic might not be enough to understand the different properties of the measured variable. Instead the distribution could be summarized using some summary statistics together, each of which can highlight different properties of the data. Tukey's five number summary, letter values or quantiles can serve as examples of distributional summaries. Another way of summarizing could be estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. Several forms of  non-parametric density estimators <!--and quantile estimators--> exist in the literature, the most commonly used form in @Silverman1986-cl<!--,  @Jones1992-pj)--> is as follows:

@@

### Interaction

>>>>

Effective exploration or good visualization require good data structures. It is recognized that there is a crucial influence of linear vs cyclic time characteristics on the expressiveness of visualization and analysis. Moreover, one can use calendars based on application domain that define contextual system of linear and cyclic granularities. However, restructuring time in this manner leads to re-organisation of the data structure, where each level of a cyclic granularity correspond to multiple values of the measured variable, unlike the case of linear granularities where each time point corresponds to an unique value of the measured variable. It is common to see consolidation of these multiple values through aggregation or an unique summary statistic like mean or median to have each level of cyclic granularity to correspond to an unique value of the measured variable. But that approach  hides the distributions of the measured variable induced by the re-organised data structure. Summarizing the probability distribution of these multiple observations using some summary statistics together, each of which can highlight different properties of the data could be a potential way to compehend periocities in the data. Hence, a possible data structure would be to consider different summarisation of the probability distribution of the measured variable across two cyclic granularities. Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might become clumsy. Instead, the big problem is broken down into smaller pieces by focusing on two cyclic granularities at a time.

<!--Since, we are interested in detection of unknown periodic behavior of data, it makes sense to support the detection of patterns by obtaining statistical summaries across cyclic time granularities.--> 
<!-- Any attempt to encode all or many cyclic granularities at the same time to develop insights on periodicity might fail or become clumsy. Instead, the big problem is broken down into smaller pieces by focusing on two cyclic granularities at a time. -->

@@

The way cyclic granularities relate become important when we consider the data structure in Figure \@ref(fig:data-structure). While considering data sets of the form $<C_1, C_2, v>$, let $S_{ij}$ be the set of index set such that for all $s \in S_{ij}$, $C_1(s) = A_i$ and $C_2(s) = B_j$. Data subsets like <$A_i$, $B_j$, $v(s)$> can be obtained for all $i \in {1, 2, \dots, n}$ and $j \in {1, 2, \dots, m}$  which will lead to $nm$ data subsets. We will discuss few cases, where one or more of these $nm$ sets will be empty, the implication of which we will discuss in Section \ref{sec:vis-interaction}.

<!--To consider a general framework, we will define pairs of cyclic granularities as harmonies or clashes.

@@

Firstly, empty combinations can arise due to the structure of the calendar or hierarchy.
These are "structurally" empty combinations in that it is impossible for them to have any value of the measured variable. Let us take a specific example, where $C_1$ be day-of-month and $C_2$ be week-of-month. Here $C_1$ can have 31 levels/categories while $C_2$ can have 5 categories. There will be $31\times 5=155$ sets $S_{ij}$ corresponding to the possible combinations of day-of-month and week-of-month. Many of these are empty. For example $S_{1,5}$, $S_{21,2}$, etc. This is also intuitive since the first day of the month can never correspond to fifth week of the month.

@@

Secondly, empty combinations can turn up due to differences in event location or duration in a calendar. These are called "event-driven" empty combinations. Let us consider $C_1$ be day-of-week and $C_2$ be WorkingDay/NonWorkingDay. Here $C_1$ can have 7 levels while $C_2$ can have 2 levels. So there are 14 sets $S_{ij}$ corresponding to the possible combinations of $C_1$ and $C_2$. While potentially all of these can be non-empty (it is possible to have a public holiday on any day-of-week), in practice many of these combinations will probably have very few observations. For example, there are few (if any) public holidays on Wednesdays or Thursdays in any given year in Melbourne, Australia.

@@

Thirdly, empty combinations can be a result of how granularities are constructed. These are called "build-based" empty combinations. Let $C_1$ be Business-days, which are days from Monday to Friday except holidays and $C_2$ is day-of-month. Then the days denoting weekends in a month would not correspond to any Business days and would have missing observations due to the way the granularities are constructed. This is different from structurally empty combinations because structure of the calendar does not lead to these missing combinations, but the construction of the granularity does.

An example when there will be no empty combinations could be where $C_1$ and $C_2$ maps index set to day-of-week and month-of-year respectively. Here $C_1$ can have 7 levels while $C_2$ can have 12 levels. So there are $12\times7=84$ sets $S_{ij}$ corresponding to the possible combinations of day-of-week and month-of-year. All of these are non-empty because every day-of-week can occur in every month.

@@

## Summarizing the distribution of a measured variable {#sec:measured-variable}

@@

Summarizing properties of the distribution of the measured variable can give valuable indication of features like central tendency, skewness, multimodality, tail behavior or variation. Tukey's five number summary, letter values or quantiles can serve as examples of distributional summaries. Another way of summarizing could be estimating the probability density function (continuous) or probability mass function (dicrete). The approach can be parametric or non-parametric. A non-parametric approach makes less rigid assumptions about the distribution of the observed data making the data speak more for itself. These resonate more with the notion of EDA which advocates exploring data for patterns and relationships without requiring prior hypotheses. Several forms of non-parametric density estimators <!--and quantile estimators--> exist in the literature, the most commonly used form in @Silverman1986-cl<!--,  @Jones1992-pj)--> is as follows:

@@

### Interaction {#sec:vis-interaction}
